Hydra Battles and Cie

### 2021-09-13 doc/chapter-powers.tex
fix errors in tex files
more snippets
Alectryon: chapter on addition chains 40% done

@@

  Please note that the following two lemmas \emph{do not require} 
the operation~\texttt{*} to be commutative.

\begin{Coqsrc}
Lemma power_commute : 
    forall x n p, x ^ n * x ^ p ==  x ^ p * x ^ n. 
(* Proof omitted *)

Lemma power_commute_with_x : 
    forall x n,  x * x ^ n == x ^ n * x.
(* Proof omitted *)

Lemma power_of_power : 
   forall x n p,  (x ^ n) ^ p == x ^ (p * n).
(* Proof omitted *)

@@

\begin{Coqsrc}
Lemma sqr_def : forall x, x ^ 2 ==  x * x.
(* Proof omitted *)

Lemma power_of_square : 
  forall x n, (x * x) ^ n ==  x ^ n * x ^ n.
(* Proof omitted *) 
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma binary_power_mult_ok :
  forall p a x,   binary_power_mult  x a p  ==  
                  a * x ^ Pos.to_nat p.
Proof.
  induction p as [q IHq | q IHq| ].
 (* Rest of proof omitted *)
 \end{Coqsrc}

 \begin{Coqsrc}
Lemma Pos_bpow_ok : 
  forall p x, Pos_bpow x p == x ^ Pos.to_nat p.
(* Proof omitted *)

Lemma N_bpow_ok : 
  forall n x,  x ^b n  == x ^ N.to_nat n.
(* Proof omitted *)
\end{Coqsrc}

\begin{Coqsrc}
Lemma N_bpow_ok_R : 
  forall n x, x ^b (N.of_nat n)   ==  x ^  n.
(* Proof omitted *)

Lemma Pos_bpow_ok_R : 
   forall p x, p <> 0 ->
                      Pos_bpow x  (Pos.of_nat p)   ==  x ^  p.
(* Proof omitted *)

End About_power.  
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma N_bpow_commute : forall x n p,  
                        x ^b n *  x ^b p ==  
                        x ^b p *  x ^b n.
Proof.
 intros x n p; repeat rewrite N_bpow_ok.
 rewrite power_commute; reflexivity.
Qed.  
\end{Coqsrc}

@@

\begin{Coqsrc}
Definition fib_pos n :=
  let (a,b) := Pos_bpow (M:= Mul2) (1,0) n in
  (a+b).

Time Compute fib_pos 153%positive.
\end{Coqsrc}

\begin{Coqanswer}
  68330027629092351019822533679447
     : N 
Finished transaction in 0.002 secs (0.002u,0.s) (successful)
\end{Coqanswer}

@@

\begin{Coqsrc}
Eval simpl in   fun (x:A) => x ^b 17.
\end{Coqsrc}

\begin{Coqanswer}
 = fun x : A =>
       x *
       (x * x * (x * x) * (x * x * (x * x)) *
        (x * x * (x * x) * (x * x * (x * x))))
     : A -> A  
\end{Coqanswer}

@@

\begin{Coqsrc}
Eval simpl in   fun x =>  x ^ 17.  
\end{Coqsrc}

\begin{Coqanswer}
= fun x : A =>
   x * (x *  (x *  (x *  (x *   (x *  (x *  (x *
    (x * (x * (x * (x * (x * (x * (x * (x * (x * one)))))
  )))))))))))  
\end{Coqanswer}

From these tests, we may infer that  representing exponentiation algorithms as \coq{}  functions hides
information about the real structure of the computations, particularly the sharing on intermediate computations.

@@

\begin{Coqsrc}
Definition pow_17  (x:A) :=
  let x2 := x * x in
  let x4 := x2 * x2 in
  let x8 := x4 * x4 in
  let x16 := x8 * x8 in
  x16 * x.
\end{Coqsrc}
\label{pow-17-let-in}

@@

\begin{Coqsrc}
Eval cbv  zeta beta delta [pow_17]  in  pow_17.
\end{Coqsrc}

\begin{Coqanswer}
 = fun x : A =>
       x * x * (x * x) * (x * x * (x * x)) *
       (x * x * (x * x) * (x * x * (x * x))) * x
     : A -> A 
\end{Coqanswer}

In the next section, we propose to use a \emph{data structure} for representing 
the computations that lead to the evaluation of some power $x^n$, where
intermediary results are explicitly named for further use in the rest of the computation.

@@

\begin{Coqsrc}
Inductive computation {A:Type}  : Type :=
| Return (a : A)
| Mult (x y : A) (k : A -> computation).    
\end{Coqsrc}
\subsubsection{Monadic notation}

@@

\begin{Coqsrc}
Notation "z '<---'  x 'times' y ';' e2 " :=
  (Mult x y  (fun z => e2))
    (right associativity, at level 60).
\end{Coqsrc}

@@

\begin{Coqsrc}
Example comp7 : computation  :=
  x <--- 2 times 2;
  y <--- x times 2;
  z <--- y times y ;
  t <--- 2 times z ;
  Return t.  
\end{Coqsrc}

@@

\begin{Coqsrc}
Definition chain := forall A:Type, A -> @computation A.   
\end{Coqsrc}

@@

For instance, Fig~\vref{fig:C87} shows a definition of the chain  of Figure~\ref{fig:chain-87-eucl}, for the exponent $87$.
Note that, like in PHOAS, bound variables associated with the 
intermediary results are \coq{} variables of type $A$.
\begin{figure}[h]
  \centering
  \begin{Coqsrc}
Example  C87 : chain :=
 fun A (x : A) =>
  x2 <--- x times x ;
  x3 <--- x2 times x ;
  x6 <--- x3 times x3 ;
  x7 <--- x6 times x ;
  x10 <--- x7 times x3 ;
  x20 <--- x10 times x10 ;
  x40 <--- x20 times x20 ;
  x80 <--- x40 times x40 ;  
  x87 <--- x80 times x7 ;
  Return x87.
 \end{Coqsrc}
  \caption{A chain for raising x to its $87$-th power}
  \label{fig:C87}
\end{figure}

@@

\label{C87-length}
\begin{Coqsrc}
Fixpoint computation_length {A} (a:A)(m : @computation A) 
  : nat :=
match m with
  | Mult _ _ k => S (computation_length a (k a))
  | _ => 0%nat
end.

Definition chain_length (c:chain) 
   := computation_length tt (c _ tt).

Compute chain_length C87.
\end{Coqsrc}
\begin{Coqanswer}
 = 9 : nat  
\end{Coqanswer}

@@

\begin{Coqsrc}
Fixpoint computation_execute  {A:Type} (op: Mult_op A) 
                              (c : computation) :=
match c with 
| Return x => x 
| Mult x y k => computation_execute op (k (x * y))
end.

Definition chain_execute (c:chain) {A} op  (a:A) :=
  computation_execute op (c A a).
\end{Coqsrc}

\begin{Coqsrc}
Definition computation_eval `{M:@EMonoid A E_op E_one E_eq}
           (c : computation) : A := computation_execute E_op c.

Definition chain_apply (c:chain) 
    {M:@EMonoid A E_op E_one E_eq} a : A :=
    computation_eval (c A a).
\end{Coqsrc}

@@

\begin{Coqsrc}
Compute  chain_apply C87 3%Z.
\end{Coqsrc}

\begin{Coqanswer}
 =  323257909929174534292273980721360271853387%Z
     : Z
\end{Coqanswer}

\begin{Coqsrc}
Compute chain_apply C87 (M:=M2N) (Build_M2 1 1 1 0)%N.
\end{Coqsrc}

\begin{Coqanswer}
  = {|
      c00 := 1100087778366101931%N;
      c01 := 679891637638612258%N;
      c10 := 679891637638612258%N;
      c11 := 420196140727489673%N |}
     : M2 N
\end{Coqanswer}

@@

\begin{Coqsrc}
Compute chain_trace C87.
\end{Coqsrc}

\begin{Coqanswer}
(1, Init)
       :: (2, Add 1 1)
          :: (3, Add 2 1)
             :: (6, Add 3 3)
                :: (7, Add 6 1)
                   :: (10, Add 7 3)
                      :: (20, Add 10 10)
                         :: (40, Add 20 20)
                            :: (80, Add 40 40) :: (87, Add 80 7) :: nil
     : list (positive * info)
\end{Coqanswer}

@@

\label{chain-correct-def}
\begin{Coqsrc}
Definition chain_correct_nat (n:nat) (c: chain) := 
  n <> 0 /\
  forall `(M:@EMonoid  A E_op E_one E_eq) (x:A), 
      chain_apply c x ==   x ^ n.

Definition chain_correct (p:positive) (c: chain) :=
  chain_correct_nat c (Pos.
to_nat p). 
\end{Coqsrc}

@@

 \begin{Coqsrc}
Definition optimal (p:positive) (c : chain) :=
 forall c', chain_correct p c' -> 
            (chain_length c <= chain_length c')%nat.
 \end{Coqsrc}

@@

\begin{Coqsrc}
Ltac slow_chain_correct_tac :=
  match goal with 
      [ |- chain_correct ?p ?c ] =>
      let A := fresh "A" in
      let op := fresh "op" in
      let one := fresh "one" in
      let eqv := fresh "eqv" in
      let M := fresh "M" in
      let x := fresh "x"
      in  split;
        [discriminate | 
         unfold c, chain_apply, computation_eval; simpl;
         intros A op one eq M x; monoid_simpl M; reflexivity]
  end.

Example C7_ok : chain_correct 7 C7.
Proof.
   slow_chain_correct_tac.
Qed.
\end{Coqsrc}

@@

\begin{Coqsrc}
Example C87_ok : chain_correct 87 C87.
Proof.
 Time  slow_chain_correct_tac. 
\end{Coqsrc}

\begin{Coqanswer}
 Finished transaction in 62.808 secs (62.677u,0.085s) (successful)   
 \end{Coqanswer}

@@

\begin{Coqanswer}
1 subgoal, subgoal 1 (ID 219)

  A : Type
  E_op : Mult_op A
  E_one : A
  E_eq : Equiv A
  M : EMonoid E_op E_one E_eq
  x : A
  ============================
   x * x * x * (x * x * x) * x * (x * x * x) *
   (x * x * x * (x * x * x) * x * (x * x * x)) *
   (x * x * x * (x * x * x) * x * (x * x * x) *
    (x * x * x * (x * x * x) * x * (x * x * x))) *
   (x * x * x * (x * x * x) * x * (x * x * x) *
    (x * x * x * (x * x * x) * x * (x * x * x)) *
    (x * x * x * (x * x * x) * x * (x * x * x) *
     (x * x * x * (x * x * x) * x * (x * x * x)))) *
   (x * x * x * (x * x * x) * x) ==
   x *
   (x *
    (x *
     (x *
      (x *
       (x *
        (x *
         (x *
          (x *
           (x *
            (x *
             (x *
              (x *
               (x * (x * (x * (x * (x * (x * (x * (x * (x * (x * 
                ..))))))))))))))))))))))
\end{Coqanswer}
\label{fig:big-goal}

This goal is solved by the following tactic composition:

\begin{Coqsrc}
monoid_simpl M; reflexivity. 
\end{Coqsrc}

@@

\begin{Coqsrc}
(** Binary trees of multiplications over A *)

Inductive Monoid_Exp (A:Type) : Type :=
 Mul_node (t t' : Monoid_Exp A) | One_node | A_node (a:A).

Arguments Mul_node {A} _ _.
Arguments One_node {A} .
Arguments A_node {A} _ .
\end{Coqsrc}

@@

Fixpoint flatten_aux {A:Type} (t fin : Monoid_Exp A) 
  : Monoid_Exp A :=
match t with Mul_node  t t' =>
              flatten_aux t (flatten_aux t' fin)
           | One_node  => fin
           |  x => Mul_node  x fin
end.

Fixpoint flatten {A:Type} (t: Monoid_Exp A) : Monoid_Exp A :=
match t with
| Mul_node t t' => flatten_aux t (flatten t')
| One_node => One_node
| X => Mul_node X One_node
end.
\end{Coqsrc}

@@

\begin{Coqsrc}
Function eval {A:Type} {op one eqv}
         (M: @EMonoid A op one eqv)
         (t: Monoid_Exp A) : A :=
 match t with 
            | Mul_node t1 t2 => (eval M t1 * eval M t2)%M
            | One_node => one
            | A_node a => a
end.
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma flatten_valid {A} `(M: @EMonoid A op one eqv):
forall t , eval M t == eval M (flatten t).
(* Proof omitted *)

Lemma flatten_valid_2 {A} `(M: @EMonoid A op one eqv):
forall t t' , eval  M (flatten t) == eval M (flatten t')  ->
     eval M t == eval M t'.
(* Proof omitted *)
\end{Coqsrc}

\subsubsection{Transforming a multiplication into a tree}
Let us now build a tool for building terms of type  \texttt{Monoid\_Exp $A$} out
of terms of type \texttt{A} containing multiplications of the form 
\Verb|(_ * _)%M| and the variable \texttt{one}. 
In fact, what we want to  define is an inverse of the function \texttt{flatten}.

@@

\begin{Coqsrc}
(** "Quote" tactic *)

Ltac model A  op one v :=
match v with 
| (?x  * ?y)%M => let r1 := model A op one x
                  with r2 := model A op one y 
                  in  constr:(@Mul_node A r1 r2)
| one => constr:(@One_node A)
| ?x => constr:(@A_node A x)
end.
\end{Coqsrc}

>>>>

\inputsnippets{Pow/NBpow}

@@

  Please note that the following lemmas \emph{do not require} 
the operation~\texttt{*} to be commutative.

\inputsnippets{Pow/powerCommute}

@@

\inputsnippets{Pow/sqrEqn, Pow/powerOfSquare}

@@

\inputsnippets{Pow/binaryPowerMultOk}

\inputsnippets{Pow/PosBpowOk, Pow/PosBpowOkR,
  Pow/NBpowOk,
  Pow/NBpowOkR}

@@

\inputsnippets{Pow/NBpowCommute}

@@

\emph{From Module~\href{../theories/html/additions.Fib2.html}{additions.Fib2}}

\inputsnippets{Fib2/fibPos}

@@

\inputsnippets{Pow/bpow17}

@@

\inputsnippets{Pow/naivePow17}

From these tests, we may infer that  representing exponentiation algorithms as \coq{}  functions hides
information about the real structure of the computations, particularly about  sharing intermediate computations.

@@

\inputsnippets{Pow/pow17LetIn}

\label{pow-17-let-in}

@@

\inputsnippets{Pow/evalPow17LetIn}

In the next section, we define a \emph{data structure} for representing 
the computations that lead to the evaluation of some power $x^n$, where
intermediary results are explicitly named for further use in the rest of the computation.

@@

\inputsnippets{Addition_Chains/computationDef}

\subsubsection{Monadic notation}

@@

\inputsnippets{Addition_Chains/monadicComputation}

@@

\inputsnippets{Addition_Chains/comp128}

@@

\inputsnippets{Addition_Chains/chainDef}

@@

For instance, Fig~\vref{fig:C87} shows a definition of the chain  of Figure~\ref{fig:chain-87-eucl}, for the exponent $87$.
Note that, like in PHOAS, bound variables associated with the 
intermediary results are \coq{} variables of type $A$.
\begin{figure}[h]
  \centering
  \inputsnippets{Addition_Chains/C87}
   \caption{A chain for raising x to its $87$-th power}
  \label{fig:C87}
\end{figure}

@@

\label{C87-length}

\inputsnippets{Addition_Chains/chainLength}

@@

\inputsnippets{Addition_Chains/chainExecute}

\inputsnippets{Addition_Chains/C87Apply}

@@

@@

\inputsnippets{Trace_exercise/traceC87}

@@

\label{chain-correct-def}

\inputsnippets{Addition_Chains/chainCorrect}

@@

\inputsnippets{Addition_Chains/optimalDef}

@@

\inputsnippets{Addition_Chains/slowTac}

@@

\inputsnippets{Addition_Chains/slowC87Correct}

@@

\inputsnippets{Addition_Chains/WhySoSlow}

@@

\inputsnippets{Addition_Chains/MonoidExp}

@@

\inputsnippets{Addition_Chains/flattenDef}

@@

\inputsnippets{Addition_Chains/evalDef}

@@

\inputsnippets{Addition_Chains/flattenValid}
\inputsnippets{Addition_Chains/flattenValid2}

\subsubsection{Transforming a multiplication into a tree}
Let us now build a tool for building terms of type  (\texttt{Monoid\_Exp $A$}) out
of terms of type \texttt{A} containing multiplications of the form 
\Verb|(_ * _)%M| and the variable \texttt{one}. 
In fact, what we want to  define is an inverse of the function \texttt{flatten}.

@@

\inputsnippets{Addition_Chains/modelTac}
